---
title: "Boxplots"
subtitle: "Stat 131A, Fall 2018, Prof. Sanchez"
output: pdf_document
fontsize: 12pt
urlcolor: blue
---

> ### Learning Objectives
>
> - Learn how to read boxplots
> - Learn about the `boxplot()` function
> - How to graph boxplots with `ggplot2`


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Introduction

Quantitative variables can be summarized using two groups of measures:
1) center, and 2) spread. Just like there are various measures of center 
(e.g. average, median, mode), we also have several measures of spread or 
variability:

- range
- interquartile range
- standard deviation (and variance)


In this tutorial we'll use the data set `mtcars` that comes in R.

```{r}
head(mtcars)
```

Let's analyze the variable `mpg` miles per gallon.

The function `summary()` produces basic summary statistics: the five-number summary, plus the mean:

```{r}
summary(mtcars$mpg)
```


A boxplot, or more formally box-and-whisker plot, is based on the five-number summary: minimum, 1st quartile, median, 3rd quartile, and maximum.

R has the built-in function `boxplot()` that allows you to make boxplots. You just need to pass it a vector, and R will graph a boxplot vertically oriented:

```{r default-boxplot, fig.align='center'}
boxplot(mtcars$mpg)
```

You can set the argument `horizontal = TRUE` to get a boxplot horizontally oriented:

```{r horizontal, fig.align='center'}
boxplot(mtcars$mpg, horizontal = TRUE)
```

\bigskip

In fact, you can actually store the output of `boxplot()`, for example:

```{r boxplotobj, results='hide', fig.keep='none'}
bb = boxplot(mtcars$mpg)
```

The object `bb` is an object of class `"boxplot"` which contains various elements:

```{r}
bb
```

The first element `stats` contains the five-number summary:

```{r}
# five number summary
bb$stats
```

```{r}
# compare to summary()
summary(mtcars$mpg)
```


### Fences

The function `boxplot()` has an argument `range`. This argument determines how far the plot whiskers extend out from the box. By default `range = 1.5`, this means that the whiskers extend to the most extreme data point which is no more than 1.5 times the interquartile range from the box. 

```{r boxplot-hp1, fig.align='center', out.width='75%'}
# default boxplot 
# (whiskers may not extend to the most extreme data points)
boxplot(mtcars$hp, horizontal = TRUE)
```

\bigskip

A value of zero (e.g. `range = 0`) causes the whiskers to extend to the data extremes.

```{r boxplot-hp2, fig.align='center', out.width='75%'}
# boxplot with unmodified whiskers
# (whiskers extend to the most extreme data points)
boxplot(mtcars$hp, horizontal = TRUE, range = 0)
```


### Formulas with `boxplot()`

An interesting feature of `boxplot()` is that you can pass R formulas. For example
the variable (column) `am` refers to the automatic transmission. This variable takes two values: 0 if a car is automatic, 1 if the transmission is manual (stick).

```{r mpg-am, fig.align='center', out.width='75%'}
# boxplots of mpg by transmission
boxplot(mpg ~ am, data = mtcars, horizontal = TRUE, las = 1)
```

\bigskip

A similar boxpot can be produced for the number of cylinders `cyl`

```{r mpg-ycl, fig.align='center', out.width='75%'}
# boxplots of mpg by cylinders
boxplot(mpg ~ cyl, data = mtcars, horizontal = TRUE, las = 1)
```
